# Analysis of the impact of severe weather events with respect to population health and the economy in the United States from 1950 to 2011.

## 24 May 2014

## Synopsis

This research set out to answer:

1. Which types of severe weather events are most harmful with respect to population health in the United States?

1. Which types of severe weather events have the greatest economic consequences in the US.

The research used Storm Data from the National Weather Service, which records severe weather events from 1950 to 2011.

Data on fatality and injury was used as a measure of impact to population health, and the cost of property and crop damage has been used as a measure of economic impact.

The research found that while acute weather events such as Tornados, Lightning, and Flash Floods were high in the [list of top causes](#top10health), also significant were prolonged events such as heat waves.

In the period from 1950 to November 2011, Tornados have caused the most fatalities and injuries. Heat is also a significant cause of death and injury, whereas water-related events such as flood and drought have a prominant impact to property and crops.

Floods were the most significant cause of damage to property and crops. The Napa Valley flood in 2006 had a particularly high cost (billions of dollars) probably as a result of the impact to high value vineyards and its impact on the wine industry.

The [Results](#results) section lists the top event types impacting both population health and the economy.

## Data Processing

The data for this study is from  the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.

The following R packages are used in this analysis.

```{r}
library(plyr)
library(ggplot2)
```

The raw data is downloaded from the NOAA website as a compressed CSV file: "repdata-data-StormData.csv.bz2". The following code reads the file. (The compressed file format is automatically unzipped.)

```{r source_data, cache=TRUE}
storm.data <- read.csv("data//repdata-data-StormData.csv.bz2", header=T, nrows=902298, stringsAsFactors=F)
```

The file contains 902,297 observations of 37 variables.

In particular, the following data has been used:
* Number of fatalities (FATALITIES) and injuries (INJURIES) for each type of event (EVTYPE).
* Estimated cost of damage to property (PROPDMG) and crops (CROPDMG) for each type of event.
* LATITUDE and LONGITUDE of events. (This is not available for all event types, such as heat waves and floods, which occur over a larger area of land.)

Event types included instances similar in meaning for the purposes of this analysis, and duplicates as a result of variation in spelling. For example, TORNADO is represented by TORNADOS, TORNADO F1, WATERSPOUT/TORNADOS, TORNDAO, etc. Significant categories of event types have been grouped together with the following code.

```{r}
storm.data[grepl("torn", storm.data$EVTYPE, ignore.case=T), ]$EVTYPE <- "TORNADO"
storm.data[grepl("flood", storm.data$EVTYPE, ignore.case=T), ]$EVTYPE <- "FLOOD"
storm.data[grepl("hurricane", storm.data$EVTYPE, ignore.case=T), ]$EVTYPE <- "HURRICANE"
storm.data[grepl("(tstm|t\\w+orm|typhoon)", storm.data$EVTYPE, ignore.case=T), ]$EVTYPE <- "THUNDERSTORM"
storm.data[grepl("wind", storm.data$EVTYPE, ignore.case=T), ]$EVTYPE <- "WINDS"
storm.data[grepl("heat", storm.data$EVTYPE, ignore.case=T), ]$EVTYPE <- "HEAT"
storm.data[grepl("fire", storm.data$EVTYPE, ignore.case=T), ]$EVTYPE <- "FIRE"
storm.data[grepl("frost", storm.data$EVTYPE, ignore.case=T), ]$EVTYPE <- "FROST"
storm.data$EVTYPE <- as.factor(storm.data$EVTYPE)
```

Key metrics are summarized by event type.

```{r cache=TRUE}
summary.fatalities <- ddply(storm.data, .(EVTYPE), summarize, total.fatalities=sum(FATALITIES))
summary.fatalities <- subset(summary.fatalities, total.fatalities > 0)
by.fatalities <- arrange(summary.fatalities, desc(total.fatalities))

summary.injuries <- ddply(storm.data, .(EVTYPE), summarize, total.injuries=sum(INJURIES))
summary.injuries <- subset(summary.injuries, total.injuries > 0)
by.injuries <- arrange(summary.injuries, desc(total.injuries))

summary.total <- ddply(storm.data, .(EVTYPE), summarize, total.fatalities=sum(FATALITIES), total.injuries=sum(INJURIES))
summary.total$total <- summary.total$total.fatalities + summary.total$total.injuries
summary.total <- subset(summary.total, total > 0)
by.total <- arrange(summary.total, desc(total))
```

Estimated costs for property and crop damage are each stored in two columns. The first is the amount, and the second is the unit such as "K" for thousands of dollars, "M" for millions, etc. If the unit column is empty or contains an unknown value, then a unit of 1 has been assumed. The following code normalizes the costs.

```{r cache=TRUE}
m <- c("H"=100, "K"=1000, "M"=1000000, "B"=1000000000, "N"=1)
damage <- storm.data
ind <- match(toupper(as.character(damage$PROPDMGEXP)), names(m), nomatch=5)
damage$property.cost <- damage$PROPDMG * m[ind]
ind <- match(toupper(as.character(damage$CROPDMGEXP)), names(m), nomatch=5)
damage$crop.cost <- damage$CROPDMG * m[ind]
damage <- subset(damage, property.cost > 0 | crop.cost > 0)
damage.summary <- ddply(damage, .(EVTYPE), summarize,
                        total.property=round(sum(property.cost) / 1000000000, 2),
                        total.crop=round(sum(crop.cost) / 1000000000, 2))
damage.summary$total.cost <- damage.summary$total.property + damage.summary$total.crop
by.total.cost <- arrange(damage.summary, desc(total.cost))
```

<a name="results"></a>
## Results

### Population health impact

In the period 1950 to November, 2011, tornados caused the most fatalities and injuries.

The top 10 event types impacting fatality and injury are:

```{r top_10_health, message=FALSE}
top10 <- by.total[1:10, ]
ggplot(top10, aes(total.fatalities, total.injuries)) +
  xlab("Number fatalities") +
  ylab("Number injuries") +
  ggtitle("Impact of major events on population health, US 1950-2011") +
  geom_point() +
  geom_text(aes(label=EVTYPE), size=4, angle=45, hjust=0, vjust=0) +
  scale_x_log10() + scale_y_log10()
```
- Axes are shown on a log[10] scale.

<a name="top10health"></a>

```{r cache=TRUE, echo=FALSE, results='asis'}
kable(head(by.total, 10))
```

Severe events such as tornados cause a great deal of damage over a short period of time and are very localized. Whereas events such as heat waves and strong wind cause more damage across a larger segment of the population. The plot below show the locality and area of impact of major event types in 2011 for which geographic data exists.

```{r us_health_events_2011, cache=TRUE, fig.width=9}
health <- storm.data
health$total <- health$FATALITIES + health$INJURIES
health$casualty.bracket <- cut(health$total, breaks=10)
us.boundary <- subset(health, LATITUDE < 4924 & LATITUDE > 2431 & LONGITUDE > 6657 & LONGITUDE <  12446)
us.boundary$date <- strptime(as.character(us.boundary$BGN_DATE), "%m/%d/%Y %H:%M:%S")
us.2011 <- us.boundary[format(us.boundary$date, "%Y") == "2011", ]

# include top 12 across all years in intersection as some top ranked event types do not have geo data points
us.top.6 <- us.2011[us.2011$EVTYPE %in% by.total$EVTYPE[1:12], ]

ggplot(us.top.6, aes(LONGITUDE, LATITUDE, alpha=casualty.bracket, size=casualty.bracket)) +
  scale_x_reverse() +
  facet_wrap(~ EVTYPE, nrow=3, ncol=2) +
  ggtitle("Locality and area of casualties from severe weather events for 2011") +
  geom_point(color="red")
```

* This plot does not include non-localized events such as heat waves.


### Economic impact

In the period 1950 to November, 2011, floods caused the most economic damage as measured by cost of damage to property and crops.

The top 10 event types with respect to total economic cost are:

```{r top_10_damage, message=FALSE}
top10 <- by.total.cost[1:10, ]
ggplot(top10, aes(reorder(EVTYPE, total.cost), total.cost)) +
  xlab("Event type") +
  ylab("Est.cost of property and crop damage ($B)") +
  ggtitle("Top 10 Events by Economic Impact, 1950 - 2011") +
  theme(axis.text.x=element_text(angle=90, hjust=1)) +
  coord_flip() +
  geom_bar(stat="identity", fill="steelblue")
```

```{r cache=TRUE, echo=FALSE, results='asis'}
kable(head(by.total.cost, 10))
```
